---
title: "Index Inversion"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

The section summarizes several kinds of index inversion for a separable index set.

# Arbitrary index inversion

Say we have a general problem, such as from code:
```{r}
for(i in 1:U1)
  for(j in 1:U2[i])
    for(k in 1:U3)
      y[i, j, k] <- foo(x[ r[i]+1, k+3, s[i, j]+2])
```
(We continue to use $i_1 \equiv$ `i`, $i_2 \equiv$ `j`, $i_3 \equiv$ `k`, etc.) The separable index sets are $(i_1, i_2)$ and $(i_3)$.  $i_1$ and $i_2$ must be inverted together for two reasons: the extent of $i_2$ depends on $i_1$, and $f_3 \equiv$ `s[i, j]+2` depends on both $i_1$ and $i_2$.  

We use the set $(i_1, i_2)$ as an example.

The Context for $(i_1, i_2)$ is $\{i_1 = 1\ldots U1; i_2 = 1\ldots U1[i_1]\}$.  *It does not include the third for-loop.*

Given any pair $(f_1, f_3)$, we seek all values of $(t_1, t_2)$ such that there is an edge from $x[f_1, f_2(i_3), f_2]$ to $y[t_1, t_2, t_3(i_3)]$, where the functions of $i_3$ are ignored since they will be inverted separately.

In a nutshell the inversion we seek can be done by unrolling the for loops and choosing some good way to manage all the declared relationships.  This is a microcosm of what we've done previously.

The steps I've used are:

- Define sizes and offsets (if they don't start at 1) for the relevant dimensions of $x$ and $y$.  Note these do not need to be the full sizes of $x$ and $y$ that may be created elsewhere in model code.  They only need to handle what is in this declaration.
- Define row-wise "flat" indices for each $x$ and $y$ based on their sizes and offsets.
- Define `iRow` as a cumulative unrolled index.  E.g. `iRow` starts at 1 and increments everytime the innermost loop increments.  `iRow` indexes all declarations made by this declaration in its Context.
- Set up a list indexed by flat index of $y$ containing the set of `iRow`s of declarations that use that element of $y$.
- Set up a list indexed by `iRow` containing every flat index of $x$ used by that declaration.

Now given indices of $y$, we can go through the following steps:

- Determine the flat indices of $y$.
- Determine which unrolled declarations used that element of $y$.
- Determine which elements of $x$ were defined by those declarations.

There are certainly other ways this could all be done.

# Block inversion

This handles a situation like:

```
for(i in 1:U)
  y[i+2] <-foo(x[i-1])
```

We handle this via an offset.

# All inversion

This handles a situation like:

```
for(i in 1:U)
  y[i] <- x
```


# Constant inversion

This handles a situation like:

```
y[2] <- x
```

# Blank inversion

This handles a situation like:

```
y <- sum(x[])
```

(This is not yet handled.)


