---
title: "Graph relationships"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction

Say we have model code:
```{r}
for(i in 1:l) {
  for(j in 1:p[i]) {
    y[i, j] ~ dnorm(x[i+1, k[j]], sd = sigma[j])
  }
}
```

This creates two sets of edges:

- from some elements of `x` to some elements of `y`; and
- from some elements of `sigma` to some elements of `y`.

The two sets of edges can be considered separately.  We will consider the edges from `x` to `y` below.

# Context

We define the *context* of a model declaration as the for-loops that enclose it.  In the above example, the context is $C = \{i,j: i=1\ldots l; j=1\ldots p[i] \}$.

# General formulation

Say there are $D$ indices in a context, denoted $I = (i_1, \ldots, i_D)$. Suppsed $y$ has $M$ dimensions and $x$ has $N$ dimensions. The general formulation of a set of edges in a context is the following:

$\newcommand{\mbt}{\mathbf{t}}$
$\newcommand{\mbf}{\mathbf{f}}$
$\newcommand{\mbT}{\mathbf{T}}$
$\newcommand{\mbF}{\mathbf{F}}$

\[
y[\mbt_1(i_1, i_2, \ldots, i_D), \ldots, \mbt_{M}(i_1, i_2, \ldots, i_D)] \leftarrow x[\mbf_{1}(i_1, i_2, \ldots, i_D), \ldots, \mbf_{N}(i_1, i_2, \ldots, i_D)]
\]
\[
\mbox{in Context } C = \{i_1,i_2: i_1=L_1\ldots U_1; i_2=L2(i_1)\ldots U_2(i_1) \}
\]

Some comments:

- For generic context (for-loop) indices:

    - We switch from $(i, j, k, \dots)$ to $(i_1, i_2, i_3, \ldots)$.
    - Index $i_1$ runs from $L_1\ldots U_1$.  Index $i_2$ runs from $L_2(i_1)\ldots U_2(i_1)$, etc.

- $\leftarrow$ indicates $x$ is upstream in the graph and $y$ is downstream, so the graph edges are directed **from** parts of x **to** parts of y.
- Indices "t" stand for "to" and "f" for "from".
- Indices are boldface to indicate they could be vectors of indices.  E.g., $\mbt_1$ could be `1:100`.  Non-boldface indices are scalars.
- In implementation, indices for the full extent of a dimension can be indicated by a blank.
- The context extent of each index variable can be independent or can depend on outer index variables.

For further index notation, define:

- $T = (t_1, \ldots, t_M)$
- $F = (f_1, \ldots, f_N)$
- $\mbT = (\mbt_1, \ldots, \mbt_M)$
- $\mbF = (\mbf_1, \ldots, \mbf_N)$

In words:

- Lowercase plain indicates a single scalar index.
- Lowercase bold indicates a block for a single index.
- Capital plain indicates a full set of scalar indices.  
- Capital bold indicates a (possible) block of indices.

## Index inversion for downstream dependents

We look at the problem of querying what elements of $y$ depend on a particular (scalar) element of $x$ as an index inversion.  Given $(f_1, \ldots, f_N)$, determine all:

\[
\{(t_1, \ldots, t_M) :  (f_1(I), \ldots, f_N(I)) = (f_1, \ldots, f_N); (t_1(I), \ldots, t_M(I)) = (t_1, \ldots, t_M); I \in C \}
\]

or in more compact form, given $F$, determine all:
\[
\{T : F(I) = F; T(I) = T; I \in C\}
\]

# Separable indices

The above formulation takes all dimensions together.  Of particular interest are cases like the following, using $M = N = 2$;

\[
y[\mbt_1(i_2), \mbt_{2}(i_1)] \leftarrow x[\mbf_{1}(i_1), \mbf_{2}(i_2)]
\]
\[
\mbox{in Context } C \mbox{ such that } L_2(i_1) = L_2, U_2(i_1) = U_2. \mbox{ (Index extents are independent.)}
\]

This would result from code such as:
```{r}
for(i in 1:U1)
  for(j in 1:U2)
    y[j, i] <- foo(x[ r[i], s[j])
```

where `r` and `s` are index vectors.  The use of `i` and `j` in different order for `y` and `x` is just an illustration to support general thinking.

In this example, the index inversion can be separated for `i` and `j`.  (`i` is $i_1$ and `j` is $i_2$.)  Given $(f_1, f_2)$, find all:

\[
\{t_2 : f_1(i_1) \equiv r[i_1] = f_1, t_2(i_1) \equiv i_1 = t_2, i_1 \in 1\ldots U_1 \}
\]

and

\[
\{t_1 : f_2(i_2) \equiv s[i_2] = f_2, t_1(i_2) \equiv i_2 = t_1, i_2 \in 1 \ldots U_2 \}
\]

# Separable index sets

Sometimes separability will only occur for multiple indices.  Consider this case:
```{r}
for(i in 1:U1) 
  for(j in 1:U2)
    for(k in 1:U3)
      y[i, j, k] <- foo(x[ r[i,j], s[i,j], q[k]])
```

Now the index set `(i, j)` (or $(i_1, i_2)$) must be inverted jointly, while `k` (or $i_3$) can be inverted on its own.  This can be seen because given $f_1 = r[i_1, i_2]$ and $f_2 = s[i_1,i_2]$, both $i_1$ and $i_2$ must satisfy the condition together to find a valid $y[i_1, i_2, i_3]$.

The same can be true in the opposite direction.  Consider this case:
```{r}
for(i in 1:U1) 
  for(j in 1:U2)
    for(k in 1:U3)
      y[i, 1:v[i, j], k] <- foo(x[ i, j, k])
```
In this case, the extent of the second dimension of $y$ depend on indices used in the first two elements of $x$.  Again, `(i, j)` must be inverted jointly.

Another case that forces two indices to be inverted jointly is when the extent of one depends on the value of the other. Consider this case:
```{r}
for(i in 1:U1) 
  for(j in 1:U2[i])
    for(k in 1:U3)
      y[i, j, k] <- foo(x[ i, j, k])
```
In this case, the extent of `j` depends on `i`, so again `(i, j)` must be inverted jointly.

### Definition of *separable index set*

We define a **separable index set** $(k_1, \ldots, k_d)$ to be the dimensions of $I$ such that some dimensions of $F$ can be inverted independent of other dimensions.  In the first example above, $(1, 2)$ is a separable index set because dimensions $(f_1, f_2)$ of $F$, which depend on $(i_{k_1} = i_1, i_{k_2} = i_2)$, can be inverted independently of $f_3$.

# Fixed index values.

Sometimes there are elements of $T$ or $F$ that are independent of $I$.  This could either be due to a constant or a blank entry (indicating a full index range).  

Consider several examples.

### Example 1:
```{r}
for(i in 1:U1)
  for(j in 1:U2)
  y[i, j] <- foo(x[i, ])
```

In this case, given $F = (f_1, f_2)$, the solution for $\mbt_2$ is always `1:U2`.

### Example 2:
```{r}
for(i in 1:U1)
  for(j in 1:U2)
  y[i, j] <- foo(x[i, 3])
```

In this case, given $F = (f_1, f_2)$, the solution for $\mbt_2$ is `1:U2` if $f_2 = 3$, and is an empty set otherwise.

### Example 3:
```{r}
for(i in 1:U1)
  for(j in 1:U2)
  y[i, ] <- foo(x[i, j])
```

This code should not really be written or allowed, although JAGS seems to allow it.  In this case, given $F = (f_1, f_2)$, the solution for $\mbt_2$ is `<blank>` (the full range, which may not be known at some stages of implementation) if $j \in$ `1:U2`, and is an empty set otherwise.

### Example 4:
```{r}
for(i in 1:U1)
  for(j in 1:U2)
  y[i, 3] <- foo(x[i, j])
```

Again, this code should not really be written or allowed.  In this case, given $F = (f_1, f_2)$, the solution for $\mbt_2$ is `3` if $j \in $ `1:J`, and is an empty set otherwise.

More cases could be considered.  What seems to emerge is that fixed index values can be handled simply as a single separable index set

# 